---
title: "Assignment_1"
author: "Alexis McCartney"
date: "2025-09-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
# Load my Data
library(readxl)
Cereals <- read_excel("~/Desktop/BA64060/Cereals.xlsx")
```

# Quantitative Variable: Calories
I grouped the calories of Cereal brands by variable.
```{r}
```
**Findings:** The Cereal brands with the fewest calories has 50 calories. 25% of cereals have 100 or fewer calories. Half of the cereals are greater or fewer than 110 calories. Average calories across most brands are 106.9 and 75% of cereals have 110 or fewer calories.

# Categorical Variable: Manufacturer
I grouped the manufacturer variable
```{r}
table(Cereals$mfr)
```
**Findings:** The majority of Cereal brands are manufactured by Kellogg & General Mills. Smaller manufacturers like A and N are not as represented in the data, as they appear once.

# Variable Transformation
I transformed Cereal names into all lowercase.
```{r}
```
Cereals$name_lower<- tolower(Cereals$name)
head(Cereals[, c("name", "name_lower")], 10)
**Findings:** The original cereal names reveal lower case versions.

# Quantitative Variable Plot
Histogram of carbohydrates (`carbo`)
```{r}
```
hist(Cereals$carbo,
main = "Carbohydrates in Cereals",
xlab = "Carbohydrates (grams)",
col = "pink",
border = "green")
**Findings:** Most Cereals are in the 10-15g carbohydrate range.

# Quantitative Variable Scatterplot
Scatterplot of **Carbohydrates versus Sodium**
```{r}
```
plot(Cereals$carbo, Cereals$sodium,
main = "Sodium vs Carbohydrates in Cereals",
xlab = "Carbohydrates (grams)",
ylab = "Sodium",
col = "pink"
pch = 16)
**Findings:** Each point represents a Cereal.The findings show if cereals with higher carbohydrates concentration also has more sodium.

